---
title: "Random Forest & Ridge Breakout/Bounce Back Predictor"
author: "Ryan J. Clark"
date: "2025-07-18"
output: html_document
---
# Using the Difference in Expected Statistics and Rate Statistics to Predict Breakout Stars with Random Forest

```{r message=FALSE, warning=FALSE}
#Loading Packages

library(tidyverse)

library(randomForest)

library(glmnet)

library(ggplot2)

library(readxl)

library(knitr)

```

### I started by downloading expected and rate statistics for 2023 and OPS for 2024
```{r}
#Importing Data 
Offense_2023 <- Offense_2023 <- read_excel("Offense_2023.xlsx", skip = 1)

Offense_2024 <-  read_excel("Offense_2024.xlsx", skip = 1)


#Standardizing names 
Offense_2023 <- Offense_2023 %>%
  separate(`last_name, first_name`, into = c("Last", "First"), sep = ", ") %>%
  mutate(Player = paste(First, Last)) %>%
  select(-First, -Last)

Offense_2024 <- Offense_2024 %>%
  separate(`last_name, first_name`, into = c("Last", "First"), sep = ", ") %>%
  mutate(Player = paste(First, Last)) %>%
  select(-First, -Last)
```

### I then selected the required columns from both datasets.
```{r}
# Select required data 

OPS_2024 <- Offense_2024 %>% select(Player, player_age, on_base_plus_slg)

Offense_2023_Clean <- Offense_2023 %>% select(Player, pa, xbadiff, xslgdiff, oz_swing_percent, whiff_percent, barrel_batted_rate)
```

### I merged the two together to have 2024 OPS in the 2023 in the stats dataset
```{r}
#Merging 


Prediction_Data_2023 <- full_join(Offense_2023_Clean, OPS_2024, by = "Player")
```

### I wanted to use players who were younger and lacking a full season of at bats.
```{r}
#Filtering Data to between 100 and 300 PAs, Age >= 27

Prediction_Data_2023_Filtered <- Prediction_Data_2023 %>% filter(player_age <= 27, pa >= 100, pa <=300) 

Prediction_Data_2023_Filtered <- drop_na(Prediction_Data_2023_Filtered)

```

### Creating the models 
```{r}
#Building the Forest 
set.seed(21)

features <- c("xslgdiff", "xbadiff", "whiff_percent", "barrel_batted_rate", "oz_swing_percent")  
x_train <- Prediction_Data_2023_Filtered[, features]
y_train <- Prediction_Data_2023_Filtered$on_base_plus_slg


rf_model <- randomForest(x = x_train, y = y_train, ntree = 500, importance = TRUE)
```

### I then added 2024 expected stats to be used for predicting 2025 OPS
```{r}
#Filtering 2024 

Prediction_Data_2024 <- Offense_2024 %>% select(Player, player_age, pa, xbadiff, xslgdiff, oz_swing_percent, barrel_batted_rate, whiff_percent)

Prediction_Data_2024_Filtered <- Prediction_Data_2024 %>% filter(player_age <= 27, pa >= 100, pa <=300) 

Prediction_Data_2024_Filtered <- drop_na(Prediction_Data_2024_Filtered)


```
### Predicting 2025 OPS with the model
```{r}

x_test <- Prediction_Data_2024_Filtered[, features]  
pred_2025 <- predict(rf_model, newdata = x_test)


Prediction_Data_2024_Filtered$Predicted_OPS_2025_RF <- pred_2025
```

### Plotting the top 10 2024 Rookies by Expected Stats
```{r}
#Plotting 

breakouts <-Prediction_Data_2024_Filtered %>% 
  arrange(desc(Predicted_OPS_2025_RF))
 
Rookies_2024 <- read_excel("Rookies_2024.xlsx", skip = 1)

rookie_breakouts <- breakouts %>%
  filter(Player %in% Rookies_2024$Name) %>% 
  distinct(Player, .keep_all = TRUE) %>% 
  slice(1:10)

ggplot(rookie_breakouts, aes(x = reorder(Player, Predicted_OPS_2025_RF), y = Predicted_OPS_2025_RF)) +
  geom_col(fill = "royalblue") +
  geom_text(aes(label = round(Predicted_OPS_2025_RF, 3)), hjust = 1) +
  coord_flip() +
  labs(title = "Top 10 Predicted Breakout 2025 Rookie Candidates",
       x = "Player", y = "Predicted OPS") +
  theme_minimal()
```

### Plotting the top 10 2024 Non-rookies by Expected Stats
```{r}
non_rookies <- breakouts %>% 
  filter(!Player %in% Rookies_2024$Name) 

non_rookies_10 <- non_rookies %>% 
  distinct(Player, .keep_all = TRUE) %>% 
  slice(1:10)


ggplot(non_rookies_10, aes(x = reorder(Player, Predicted_OPS_2025_RF), y = Predicted_OPS_2025_RF)) +
  geom_col(fill = "purple") +
  geom_text(aes(label = round(Predicted_OPS_2025_RF, 3)), hjust = 1) +
  coord_flip() +
  labs(title = "Top 10 Predicted 2025 Bounce Back Candidates",
       x = "Player", y = "Predicted OPS") +
  theme_minimal()
```

### Feature Importance 
```{r}

kable(importance(rf_model))
varImpPlot(rf_model)




```

# Now for Ridge Regression 


### Model 
```{r}

#Building the Model 
set.seed(21)

x_2023 <- model.matrix(on_base_plus_slg ~ xslgdiff + xbadiff + whiff_percent + barrel_batted_rate + oz_swing_percent, Prediction_Data_2023_Filtered)[, -1]

y_2024 <- Prediction_Data_2023_Filtered$on_base_plus_slg

cv_ridge <- cv.glmnet(x_2023, y_2024, alpha = 0, nfolds = 10)
best_lambda <- cv_ridge$lambda.min
```

### Using the Ridge model to predict 2025 
```{r}
#Predicting 2025 OPS with Ridge 
x_2024 <- model.matrix(~ xslgdiff + xbadiff + whiff_percent + barrel_batted_rate + oz_swing_percent, Prediction_Data_2024_Filtered)[, -1]

pred_2025 <- predict(cv_ridge, s = best_lambda, newx = x_2024)

Prediction_Data_2024_Filtered$Predicted_OPS_2025_Ridge <- as.vector(pred_2025)
```
### Plotting ridge regression breakout/ bounceback candidates 

```{r}
breakouts <-Prediction_Data_2024_Filtered %>% 
  arrange(desc(Predicted_OPS_2025_Ridge))
 
Rookies_2024 <- read_excel("Rookies_2024.xlsx", skip = 1)

rookie_breakouts <- breakouts %>%
  filter(Player %in% Rookies_2024$Name) %>% 
  distinct(Player, .keep_all = TRUE) %>% 
  slice(1:10)

ggplot(rookie_breakouts, aes(x = reorder(Player, Predicted_OPS_2025_Ridge), y = Predicted_OPS_2025_Ridge)) +
  geom_col(fill = "royalblue") +
  geom_text(aes(label = round(Predicted_OPS_2025_Ridge, 3)), hjust = 1) +
  coord_flip() +
  labs(title = "Top 10 Predicted Breakout 2025 Rookie Candidates Ridge",
       x = "Player", y = "Predicted OPS") +
  theme_minimal()
```
```{r}
non_rookies <- breakouts %>% 
  filter(!Player %in% Rookies_2024$Name) 

non_rookies_10 <- non_rookies %>% 
  distinct(Player, .keep_all = TRUE) %>% 
  slice(1:10)


ggplot(non_rookies_10, aes(x = reorder(Player, Predicted_OPS_2025_Ridge), y = Predicted_OPS_2025_Ridge)) +
  geom_col(fill = "purple") +
  geom_text(aes(label = round(Predicted_OPS_2025_Ridge, 3)), hjust = 1) +
  coord_flip() +
  labs(title = "Top 10 Predicted 2025 Bounce Back Candidates",
       x = "Player", y = "Predicted OPS") +
  theme_minimal()
```



### Backtraining model 
```{r}

pred_2024 <- predict(cv_ridge, s = best_lambda, newx = x_2023)

# RMSE
rmse <- sqrt(mean((pred_2024 - y_2024)^2))

# R²
r2 <- cor(pred_2024, y_2024)^2

# MAE
mae <- mean(abs(pred_2024 - y_2024))

#Making a table 
model_accuracy <- data.frame(
  Metric = c("RMSE", "R²", "MAE", "Lambda"),
  Value = c(rmse, r2, mae, best_lambda)
)

kable(model_accuracy)
```

